Formalizing the Edmonds-Karp Algorithm

### 2016-03-04 paper/root.tex
Reza's mods

\maketitle
\begin{abstract}
We present a formalization of the Ford-Fulkerson method for computing the maximum flow in a network.
Our formal proof closely follows a standard textbook proof, and is easy to read even without being
an expert in Isabelle/HOL--- the interactive theorem prover used for the formalization.
% 
% 
% so it is easy to read even for those who 
% are not expert in Isabelle/HOL--- the interactive theorem prover used for the formalization.

We then use stepwise refinement to obtain the Edmonds-Karp algorithm, and formally prove a bound on its complexity.
Further refinement yields a verified implementation, which compares well to an unverified reference implementation in Java.

@@

\section{Introduction}
The maximum flow problem is an important problem in graph theory, as many other problems like the maximum-bipartite-matching problem, the edge-disjoint-paths problem, the circulation-demand problem, and various scheduling and resource allocating problems can be reduced to it.
The Ford-Fulkerson method~\cite{FF56} describes a class of algorithms to solve the maximum flow problem, an important instance being the Edmonds-Karp algorithm~\cite{EK72},
which solves the maximum flow problem in polynomial time for the general case of real valued capacities.

@@

An flow $f$ on a network is a labeling of the edges with real values satisfying the following constraints: 1) \emph{Capacity constraint}: the flow on each edge is a non-negative value smaller or equal to the edge's capacity; 2) \emph{Conservation constraint}: For all vertices except $s$ and $t$, the sum of flows over all incoming edges is equal to the sum of flows over all outgoing edges.

The value of flow $f$ is denoted by $|f|$, and defined to be the sum over the flows of all outgoing edges of the source\footnote{Note that we assumed that the source has no incoming edges. In a more general setting, the value of a flow is the sum over all flows leaving $s$ minus the sum over all incoming flows to $s$.}.
Given a network $G$, the maximum flow problem is to find a flow with a maximum value among all flows of the network.

@@

The Ford-Fulkerson method is a corollary of this theorem. It is based on a greedy approach: Starting from a zero flow, the value of the flow is iteratively increased until a maximum flow is reached. In order to increase the flow value, it may be necessary to redirect some flow, \ie to decrease the flow passed through specific edges. For this purpose, the Ford-Fulkerson method defines the residual graph. It has edges in parallel and reverse to the edges of the network, each edge indicating how much flow can be moved between two nodes by either increasing or decreasing the flow on an edge in the network. Formally, the residual graph $G_f$ of a flow $f$ is the graph induced by the 
edges with positive labels according to the following labeling function $c_f$:
\[ c_f (u, v) = 
  \begin{cases}
  c (u, v) - f(u, v) \hfill & \text{ if $(u, v) \in E$} \\
  f (v, u) \hfill & \text{ if $(v, u) \in E$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]

@@

TODO: Introduce obtain-syntax, and unfold find-augmenting-path spec!
The code looks quite similar to pseudocode that one would expect in a textbook, but actually is a rigorous formal specification of the algorithm, using nondeterminism to leave open the implementation details (\cf Section~\ref{sec:refinement}). The only caveat is that our framework currently does not support breaking of loops, such that we had to use an explicit flag instead.

@@

  We conclude with some statistics:
  The formalization consists of roughly 8000 lines of proof text, where the graph theory up to the Ford-Fulkerson theorem required 2500 lines.
  The abstract Edmonds-Karp algorithm and its complexity analysis contributes 900 lines, and its implementation (including BFS) another 1500 lines.
  The remaining lines are contributed by the network checker and some auxiliary theories. The development of the theories required roughly 3 man month, a significant amount of this time going into a first, purely functional version of the implementation, which was later dropped in favor of the faster imperative version.
  TODO: Reza, is this estimate OK?

  \subsection{Related Work}
  We are only aware of one other formalization of the Ford-Fulkerson method conducted in Mizar~\cite{MaRu05} by Lee. Unfortunately, there seems to be no publication
  on this formalization except~\cite{Lee05}, which provides a Mizar proof script without any additional comments except that it ``defines and proves correctness of Ford/Fulkerson's Maximum Network-Flow algorithm at the level of graph manipulations.''. Moreover, in Lee et al.~\cite{LeRu07}, which is about graph representation in Mizar, the formalization is shortly mentioned, and it is clarified that it does not provide any implementation or data structure formalization.
  As far as we could understand the Mizar proof scripts, it provides an abstract formalization of the Edmonds/Karp algorithm, including breadth first search. Termination is only proved for the case of integer valued capacities, and no implementation is provided.

>>>>

\maketitle
\begin{abstract}
We present a formalization of the Ford-Fulkerson method for computing the maximum flow in a network.
Our formal proof closely follows a standard textbook proof, and is easy to read even without being
an expert in Isabelle/HOL--- the interactive theorem prover used for the formalization.
We then use stepwise refinement to obtain the Edmonds-Karp algorithm, and formally prove a bound on its complexity.
Further refinement yields a verified implementation, which compares well to an unverified reference implementation in Java.

@@

\section{Introduction}
Computing the maximum flow of a network is an important problem in graph theory.
Many other problems, like maximum-bipartite-matching, edge-disjoint-paths, circulation-demand, and various scheduling and resource allocating problems can be reduced to it.
The Ford-Fulkerson method~\cite{FF56} describes a class of algorithms to solve the maximum flow problem, an important instance being the Edmonds-Karp algorithm~\cite{EK72},
which solves the maximum flow problem in polynomial time for the general case of real valued capacities.

@@

An \emph{$s$-$t$ flow} $f$ on a graph is a labeling of the edges with real values satisfying the following constraints: 1) \emph{Capacity constraint}: the flow on each edge is a non-negative value smaller or equal to the edge's capacity; 2) \emph{Conservation constraint}: For all vertices except $s$ and $t$, the sum of flows over all incoming edges is equal to the sum of flows over all outgoing edges. In context of a network, a \emph{flow} is an $s$-$t$ flow, \ie a flow from the source to the sink.

The value of an $s$-$t$ flow $f$ is denoted by $|f|$, and defined to be the sum over all flows leaving $s$ minus the sum over all incoming flows to $s$\footnote{In the context of networks, there are no incoming edges to $s$, such the definition can be simplified.}.
Given a network $G$, the maximum flow problem is to find a flow with a maximum value among all flows of the network.

@@

The Ford-Fulkerson method is a corollary of this theorem. It is based on a greedy approach: Starting from a zero flow, the value of the flow is iteratively increased until a maximum flow is reached. In order to increase the flow value, it may be necessary to redirect some flow, \ie to decrease the flow passed through specific edges. For this purpose the Ford-Fulkerson method defines the residual graph, which has edges in parallel and reverse to the edges of the network. 
Each edge indicates how much flow can be moved between two nodes by either increasing or decreasing the flow on an edge in the network. Formally, the residual graph $G_f$ of a flow $f$ is the graph induced by the 
edges with positive labels according to the following labeling function $c_f$:
\[ c_f (u, v) = 
  \begin{cases}
  c (u, v) - f(u, v) \hfill & \text{ if $(u, v) \in E$} \\
  f (v, u) \hfill & \text{ if $(v, u) \in E$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]

@@

TODO: Introduce obtain-syntax, and unfold find-augmenting-path spec! Or even while-exists loop as combinator?
The code looks quite similar to pseudocode that one would expect in a textbook, but actually is a rigorous formal specification of the algorithm, using nondeterminism to leave open the implementation details (\cf Section~\ref{sec:refinement}). Note that we had to use the available combinators of the refinement framework, which made the 
code slightly more verbose than we would have liked.

% The only caveat is that our framework currently does not support breaking of loops, such that we had to use an explicit flag instead.

@@

  We conclude with some statistics:
  The formalization consists of roughly 8000 lines of proof text, where the graph theory up to the Ford-Fulkerson theorem required 2500 lines.
  The abstract Edmonds-Karp algorithm and its complexity analysis contributes 900 lines, and its implementation (including BFS) another 1500 lines.
  The remaining lines are contributed by the network checker and some auxiliary theories. The development of the theories required roughly 3 man month, a significant amount of this time going into a first, purely functional version of the implementation, which was later dropped in favor of the faster imperative version.

  \subsection{Related Work}
  We are only aware of one other formalization of the Ford-Fulkerson method conducted in Mizar~\cite{MaRu05} by Lee. Unfortunately, there seems to be no publication
  on this formalization except~\cite{Lee05}, which provides a Mizar proof script without any additional comments except that it ``defines and proves correctness of Ford/Fulkerson's Maximum Network-Flow algorithm at the level of graph manipulations.''. Moreover, in Lee et al.~\cite{LeRu07}, which is about graph representation in Mizar, the formalization is shortly mentioned, and it is clarified that it does not provide any implementation or data structure formalization.
  As far as we could understand the Mizar proof scripts, it formalizes an algorithm roughly equivalent to our abstract version of the Ford-Fulkerson method.
  Termination is only proved for the case of integer valued capacities, and no implementation is provided.
